---
title: |
  | \vspace{8cm} \textbf{FDA Project}
author:
- Bassel MASRI
- Cyril DEVEDEUX
date: "2/14/2021"
output:
  pdf_document:
    number_sections: yes
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction

# Defining the data and the problem

Moisture (fds): This data set consists of near-infrared reflectance spectra of 100 wheat samples, measured in 2 nm intervals from 1100 to 2500nm, and an associated response variables, the samples’ moisture content.

# Approach and results

## Exploratory data analysis

Include FPCA

## Modeling

Modeling without FPCA
modeling with FPCA

# Conclusion


```{r, include=FALSE}
#-- load libraries
rm(list=ls())
library(fda)
library(refund)
library(tidyverse)
if (!require(fds)) install.packages("fds")
```

```{r, fig.width=12, fig.height=5}
#-- load data
data("Moisturespectrum")

#-- load response variable
data("Moisturevalues")
Y = Moisturevalues
X = Moisturespectrum

#-- plot the data
par(mfrow=c(1,2))
plot(X, col = "grey")
plot(Y, xlab="Sample",ylab="Moisture content", pch=15)
```

```{r, fig.width=12, fig.height=5}
mat = X$y

N = dim(mat)[1]
M = dim(mat)[2]

#-- set number of basis and generate it
basis = create.bspline.basis(c(1,M), nbasis = 25, norder = 4)

#-- plot the generated basis
mat_fd = Data2fd(1:M,t(mat),basis)
plot(mat_fd, lty = 1, col = "grey")
```

```{r}
#-- Ploting the first 30 smoothed trajectories with point-wise mean and sd
plot(mat_fd[1,], ylim = c(0.2,0.5))
```

TO DO EDA

Point wise mean
Point wise std
On the same plot, show upper bound and lower bound and pointwise mean along with the smoothed data.
Smoothing with GCV , show plot of lambda, smooth the data with best lambda, plot the curvres again
Repeat mean, std and confidence intervals
FPCA, plot the first 4 PCA's, see how much components we need to get 90% variance.

TO DO Modeling
Maybe a test train split ? (bonus)
Scalar on function regression
Evaluate model R2

Maybe scalar on function regression using FPCA (for later)

